### 2024-09-23 paper/paper.qmd
Initial commit

---
title: "TTC delays: when they happen and how to be punctual"
subtitle: "Insight into when delays disproportionately tend to happen and related statistics"
author: 
  - Andrew Goh
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "TBA"
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

```

# Introduction

You can and should cross-reference sections and sub-sections. We use @talia, @opendatatorotno, and @rohan.

The remainder of this paper is structured as follows. @sec-mydatasection

# Data {#sec-mydatasection}

Some of our data is of penguins (@fig-marriage), from @palmerpenguins.
```{r}
#| label: fig-times-mean
#| fig-cap: Subway delays w.r.t. time of day
#| echo: false

library(dplyr)
library(lubridate)
library(here)

# Load the data
subwayData <- read.csv(here::here("./data/raw_data/raw_data.csv"))
busData <- read.csv(here::here("./data/raw_data/raw_data_bus.csv"))

# Convert the Time column to a proper time format
subwayData$Time <- hm(subwayData$Time)  # Parse 'Time' as hours and minutes
busData$Time <- hm(busData$Time)

# Extract the hour from the Time column
subwayData$Hour <- hour(subwayData$Time)
busData$Hour <- hour(busData$Time)

# Aggregate data by hour to calculate total and mean delays
hourly_delays_subway <- subwayData %>%
  group_by(Hour) %>%
  summarise(Total_Delay_Subway = sum(Min.Delay, na.rm = TRUE), 
            Mean_Delay_Subway = mean(Min.Delay, na.rm = TRUE))

hourly_delays_bus <- busData %>%
  group_by(Hour) %>%
  summarise(Total_Delay_Bus = sum(Min.Delay, na.rm = TRUE), 
            Mean_Delay_Bus = mean(Min.Delay, na.rm = TRUE))

# Combine subway and bus data for comparison
hourly_delays_combined <- hourly_delays_subway %>%
  full_join(hourly_delays_bus, by = "Hour")

# Reshape data for average delays
hourly_avg_combined_long <- hourly_delays_combined %>%
  pivot_longer(cols = starts_with("Mean"), 
               names_to = "Type", 
               values_to = "Mean_Delay") %>%
  mutate(Type = recode(Type, 
                       "Mean_Delay_Subway" = "Subway", 
                       "Mean_Delay_Bus" = "Bus"))

# Plot 1: Average delays by hour of day (combined subway and bus)
ggplot(hourly_avg_combined_long, aes(x = factor(Hour), y = Mean_Delay, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Delays by Hour of Day (Subway vs Bus)", 
       x = "Hour of Day", 
       y = "Average Delay (Minutes)", 
       fill = "Type") +
  scale_fill_manual(values = c("Subway" = "steelblue", "Bus" = "tomato")) +
  theme_minimal() +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, face = "bold"))
```
```{r}
#| label: fig-times-total
#| fig-cap: total delay times
#| echo: false

# Reshape data for total delays
hourly_delays_subway <- subwayData %>%
  group_by(Hour) %>%
  summarise(Total_Delay_Subway = sum(Min.Delay, na.rm = TRUE))

ggplot(hourly_delays_subway, aes(x = factor(Hour), y = Total_Delay_Subway)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total Subway Delays by Hour of Day", 
       x = "Hour of Day", 
       y = "Total Delay (Minutes)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
```{r}
#| label: fig-times-adjustedd
#| fig-cap: delay time adjusted for the approximate subway load
#| echo: false
adjust_subway_delays <- function(data) {
  data %>%
    mutate(Adjusted_Delay = ifelse(Hour %in% 6:9 | Hour %in% 15:19, Min.Delay / 2, Min.Delay))
}

# Apply the adjustment to subway data
subwayData <- adjust_subway_delays(subwayData)

# Aggregate the adjusted data by hour for total delays
hourly_delays_subway <- subwayData %>%
  group_by(Hour) %>%
  summarise(Total_Delay_Subway = sum(Adjusted_Delay, na.rm = TRUE))

# Plot: Adjusted total subway delays by hour of day
ggplot(hourly_delays_subway, aes(x = factor(Hour), y = Total_Delay_Subway)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total Subway Delays by Hour of Day Adjusted for Load Differences", 
       x = "Hour of Day", 
       y = "Delay (Minutes)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```
Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

\newpage

# References

### 2024-09-23 scripts/01-download_data.R

@@ -1,0 +1,31 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from Open Data Toronto
+# Author: Rohan Alexander
+# Date: 19 September 2024
+# Contact: rohan.alexander@utoronto.ca
+# License: MIT
+# Pre-requisites: None
+# Any other information needed? None
+
+
+#### Workspace setup ####
+
+library(opendatatoronto)
+library(dplyr)
+
+# get package
+package <- show_package("996cfe8d-fb35-40ce-b569-698d51fc683b")
+
+# get all resources for this package
+resources <- list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b")
+
+# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
+datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+
+# load the first datastore resource as a sample
+data <- filter(resources, row_number()==10) %>% 
+  get_resource()
+
+
+#### Save data ####
+write_csv(data, "data/raw_data/raw_data.csv") 

### 2024-09-23 paper/paper.qmd
all

@@
   - Andrew Goh
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: LINK. @talia, @thereferencecanbewhatever, @dplyr were used in this paper."
 date: today
 date-format: long
-abstract: "TBA"
+abstract: "This paper uses data from Open Data Toronto (@opendatatoronto) on TTC subway and bus
+delays to analyze how the delays are generally distributed w.r.t. the time of day. Mean, total, and adjusted
+statistics are graphed and discussed to reach a conclusion."
 format: pdf
@@
 
+# Introduction
+
+The TTC, as Toronto's primary public transport network, is the means for many
+to get about in the city. At an annual cost of ~$1.4k CAD, it remains the vastly more 
+economical means of transport in Toronto, especially compared to driving costs
+averaging $13k-16k per year, according to a CAA estimate by the @ttc_know_your_ttc.
+
+With its network of subways, the average TTC rider will be, comparatively, less affected by traffic as
+the average driver is, but all is not always smooth sailing with the TTC either. Delays, whether
+it be in the bus or subway systems, are decently common in the TTC family of transportation devices,
+with each delay potentially ruining an hour's worth of brisk preparation and a day's worth of planning.
+
+While everyone has their own understanding of delays, with discrepancies amplified by personal
+experience and survivor's bias, we look into the data to determine just when and how delays happen.
+
+# Data {#sec-mydatasection}
+
+Armed with the simple driving question of how I can reduce delays in my day-to-day transits,
+we first graph the mean subway and bus delay data (@fig-times-mean) to have a broad understanding of the data.
 ```{r}
-#| include: false
+#| label: fig-times-mean
+#| fig-cap: Subway delays w.r.t. time of day
+#| echo: false
 #| warning: false
@@
 library(tidyverse)
-
-data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-```
-
-# Introduction
-
-You can and should cross-reference sections and sub-sections. We use @talia, @opendatatorotno, and @rohan.
-
-The remainder of this paper is structured as follows. @sec-mydatasection
-
-# Data {#sec-mydatasection}
-
-Some of our data is of penguins (@fig-marriage), from @palmerpenguins.
-```{r}
-#| label: fig-times-mean
-#| fig-cap: Subway delays w.r.t. time of day
-#| echo: false
-
 library(dplyr)
@@
 ```
+Two main takeaways from the figure: one, that the magnitude of delays is generally the same with each other, and two, that the average
+delay time for buses is around 4-5 times more than that of subways.
+This confirms the intuition that subways generally are a more time-reliable way to get somewhere than buses, and we will therefore focus on subways from hereonwards in our quest for the most reliable transit plan.
+
+While we don't know the total frequency of subways and therefore cannot calculate an "average delay per ride" statistic, what we can do is sum up the delay at each hour of day to still obtain an estimator for how much delay you're likely to encounter in a trip during each respective hour, as shown in @fig-times-total.
+
 ```{r}
@@
 ```
+From the data here, it seems that 8-9am and 3-5pm seem to be the worst offenders, which would suggest that
+transit during these times is the most prone to delays.
+According to the @ttc_line_1_schedule, though, subways run at approximately double the frequency during these rush hours, accounting for the increase in total delay times by sheer volume.
+Finally, adjusting for this by halving the relative delay times for the aforementioned rush hours (6-9am, 3-7pm), we obtain @fig-times-adjusted.
+
 ```{r}
-#| label: fig-times-adjustedd
-#| fig-cap: delay time adjusted for the approximate subway load
+#| label: fig-times-adjusted
+#| fig-cap: Delay time adjusted for the approximate subway load
 #| echo: false
@@
   geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Total Subway Delays by Hour of Day Adjusted for Load Differences", 
+  labs(title = "Total Subway Delays by Hour of Day Adjusted for Capacity", 
        x = "Hour of Day", 
@@
 ```
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
 
@@
 
-## First discussion point {#sec-first-point}
+Surprisingly, peak hours are the most consistent hours of them all, at least when compared to the off-hours of the TTC.
+A potential reason for this could be the higher-stress situation leading to tighter schedules, the lack of shift changes, or even the general alertness of drivers, among others - this, however, is not the focus nor concern of our current investigation.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+As to the average delay amount, given that a delay happens, we look at average ridership data from @metrolinx_rtp and @average_distance, stating that "local" trips take around 4.2km on average.
+Average distance between stations, calculated using data from @ttc_conventional_system_2023, is 70 stations / 70.1km for approximately 1km per station (on average) and a resulting average of around 4 stations per trip.
 
-## Second discussion point
-
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
-
-\newpage
-
-\appendix
-
-# Appendix {-}
-
-# Additional data details
-
+Assuming a carry-over delay for your specific subway arriving at each of the 4 stations, we conclude that for the average trip during peak hour, leaving 10-15 minutes early is a fairly safe margin to ensure punctual arrival even in the case of delays.
 \newpage

### 2024-09-23 scripts/01-download_data.R

@@ -12,6 +12,7 @@
 
 library(opendatatoronto)
 library(dplyr)
+library(tidyverse)
 
 # get package
 package <- show_package("996cfe8d-fb35-40ce-b569-698d51fc683b")
@@ -29,3 +30,11 @@ data <- filter(resources, row_number()==10) %>%
 
 #### Save data ####
 write_csv(data, "data/raw_data/raw_data.csv") 
+
+
+# get package
+package <- show_package("e271cdae-8788-4980-96ce-6a5c95bc6618")
+resources <- list_package_resources("e271cdae-8788-4980-96ce-6a5c95bc6618")
+data <- filter(resources, row_number()==10) %>% 
+  get_resource()
+write_csv(data, "data/raw_data/raw_data_bus.csv") 

### 2024-09-23 paper/paper.qmd
update cleaned data

@@
 
-# Load the data
-subwayData <- read.csv(here::here("./data/raw_data/raw_data.csv"))
-busData <- read.csv(here::here("./data/raw_data/raw_data_bus.csv"))
-
-# Convert the Time column to a proper time format
-subwayData$Time <- hm(subwayData$Time)  # Parse 'Time' as hours and minutes
-busData$Time <- hm(busData$Time)
-
-# Extract the hour from the Time column
-subwayData$Hour <- hour(subwayData$Time)
-busData$Hour <- hour(busData$Time)
+# Load the cleaned data
+subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
+busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
 
@@
 ```
-Two main takeaways from the figure: one, that the magnitude of delays is generally the same with each other, and two, that the average
-delay time for buses is around 4-5 times more than that of subways.
+Two main takeaways from the figure: one, that the average magnitude of delays is generally the same over the day, whether it be for subways or buses, and two, that the average delay time for buses is around 4-5 times more than that of subways.
 This confirms the intuition that subways generally are a more time-reliable way to get somewhere than buses, and we will therefore focus on subways from hereonwards in our quest for the most reliable transit plan.
@@
 While we don't know the total frequency of subways and therefore cannot calculate an "average delay per ride" statistic, what we can do is sum up the delay at each hour of day to still obtain an estimator for how much delay you're likely to encounter in a trip during each respective hour, as shown in @fig-times-total.
-
 ```{r}
@@
 #| echo: false
+
+library(tidyverse)
+library(dplyr)
+library(lubridate)
+library(here)
+
+# Load the cleaned data
+subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
+busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
 
@@
 #| echo: false
+
+library(tidyverse)
+library(dplyr)
+library(lubridate)
+library(here)
+
+# Load the cleaned data
+subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
+busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
+
+# Function to adjust subway delays
 adjust_subway_delays <- function(data) {

### 2024-09-23 scripts/02-data_cleaning.R

@@ -7,20 +7,24 @@
 # Pre-requisites: Need to have downloaded the data
 # Any other information needed? None.
 
-#### Workspace setup ####
-library(tidyverse)
+library(here)
+library(lubridate)
 
-#### Clean data ####
-raw_data <- read_csv("data/raw_data/raw_data.csv")
+# Read the raw data
+subwayData <- read.csv(here::here("./data/raw_data/raw_data.csv"))
+busData <- read.csv(here::here("./data/raw_data/raw_data_bus.csv"))
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |> 
-  separate(col = time_period,
-            into = c("year", "month"),
-            sep = "-") |> 
-  mutate(date = lubridate::ymd(paste(year, month, "01", sep = "-"))
-         )
-  
-#### Save data ####
-write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")
+# Convert the Time column to a proper time format
+subwayData$Time <- hm(subwayData$Time)  # Parse 'Time' as hours and minutes
+busData$Time <- hm(busData$Time)
+
+# Extract the hour from the Time column
+subwayData$Hour <- hour(subwayData$Time)
+busData$Hour <- hour(busData$Time)
+
+# Create the directory for analysis data if it doesn't exist
+dir.create(here::here("./data/analysis_data"), showWarnings = FALSE)
+
+# Write the cleaned data to new CSV files
+write.csv(subwayData, here::here("./data/analysis_data/analysis_data.csv"), row.names = FALSE)
+write.csv(busData, here::here("./data/analysis_data/analysis_data_bus.csv"), row.names = FALSE)

### 2024-09-27 scripts/00-simulate_data.R
updated scripts, added comments, and ran them to update the data

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Simulates data
-# Author: Rohan Alexander
+# Author: Andrew Goh
 # Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Contact: andrew.goh@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
 # Any other information needed? None
@@ -13,28 +13,20 @@ library(tidyverse)
 
 
 #### Simulate data ####
-set.seed(304)
+set.seed(6737151)
 
-# Define the start and end date
-start_date <- as.Date("2018-01-01")
-end_date <- as.Date("2023-12-31")
+n <- 100
+# simulate 100 dates
+dates <- seq.Date(from = as.Date("2022-01-01"), by = "day", length.out = n)
+# same for times
+times <- format(
+  as.POSIXct("2022-01-01 00:00:00") + runif(n, min = 0, max = 24 * 3600),
+  format = "%H:%M:%S"
+)
+# delay in minutes
+min_delay <- abs(round(rnorm(n, 15, 5), 0))
+# combine
+simulated_data <- data.frame(Date = dates, Time = times, Min.Delay = min_delay)
 
-# Set the number of random dates you want to generate
-number_of_dates <- 100
-
-data <-
-  tibble(
-    dates = as.Date(
-      runif(
-        n = number_of_dates,
-        min = as.numeric(start_date),
-        max = as.numeric(end_date)
-      ),
-      origin = "1970-01-01"
-    ),
-    number_of_marriage = rpois(n = number_of_dates, lambda = 15)
-  )
-
-
-#### Write_csv
-write_csv(data, file = "data/raw_data/simulated.csv")
+# write to file
+write_csv(simulated_data, file = "data/raw_data/simulated.csv")

### 2024-09-27 scripts/01-download_data.R

@@ -1,18 +1,17 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Open Data Toronto
-# Author: Rohan Alexander
-# Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Author: Andrew Goh
+# Date: 27 September 2024
+# Contact: andrew.goh@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
 # Any other information needed? None
 
-
-#### Workspace setup ####
-
 library(opendatatoronto)
 library(dplyr)
 library(tidyverse)
+
+#---------------subway data:----------------#
 
 # get package
 package <- show_package("996cfe8d-fb35-40ce-b569-698d51fc683b")
@@ -21,20 +20,28 @@ package <- show_package("996cfe8d-fb35-40ce-b569-698d51fc683b")
 resources <- list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
-
-# load the first datastore resource as a sample
-data <- filter(resources, row_number()==10) %>% 
+datastore_resources <- filter(resources, tolower(format) %in% c("csv", "geojson"))
+data <- filter(resources, row_number() == 9) %>%
   get_resource()
+data2 <- filter(resources, row_number() == 10) %>%
+  get_resource()
+# merge 2022 and 2023 data into one set
+data <- rbind(data, data2)
+# save into csv
+write_csv(data, "data/raw_data/raw_data.csv")
 
+#---------------bus data:----------------#
 
-#### Save data ####
-write_csv(data, "data/raw_data/raw_data.csv") 
-
-
-# get package
+# do the same for bus data: get data
 package <- show_package("e271cdae-8788-4980-96ce-6a5c95bc6618")
+# identify resources
 resources <- list_package_resources("e271cdae-8788-4980-96ce-6a5c95bc6618")
-data <- filter(resources, row_number()==10) %>% 
+# get data
+data <- filter(resources, row_number() == 10) %>%
   get_resource()
-write_csv(data, "data/raw_data/raw_data_bus.csv") 
+data2 <- filter(resources, row_number() == 11) %>%
+  get_resource()
+# merge
+data <- rbind(data, data2)
+# save
+write_csv(data, "data/raw_data/raw_data_bus.csv")

### 2024-09-27 scripts/02-data_cleaning.R

@@ -1,26 +1,42 @@
 #### Preamble ####
-# Purpose: Cleans the raw marriage data into an analysis dataset
-# Author: Rohan Alexander
+# Purpose: Cleans the raw delay data into an analysis dataset
+# Author: Andrew Goh
 # Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Contact: andrew.goh@utoronto.ca
 # License: MIT
 # Pre-requisites: Need to have downloaded the data
 # Any other information needed? None.
 
 library(here)
 library(lubridate)
-
+library(dplyr)
 # Read the raw data
 subwayData <- read.csv(here::here("./data/raw_data/raw_data.csv"))
 busData <- read.csv(here::here("./data/raw_data/raw_data_bus.csv"))
 
 # Convert the Time column to a proper time format
-subwayData$Time <- hm(subwayData$Time)  # Parse 'Time' as hours and minutes
+subwayData$Time <- hm(subwayData$Time) # Parse 'Time' as hours and minutes
 busData$Time <- hm(busData$Time)
 
 # Extract the hour from the Time column
 subwayData$Hour <- hour(subwayData$Time)
 busData$Hour <- hour(busData$Time)
+
+subwayData <- subwayData %>%
+  select(Time, Date, Min.Delay, Hour) %>%
+  na.omit() # Remove rows with NA values
+
+busData <- busData %>%
+  select(Time, Date, Min.Delay, Hour) %>%
+  na.omit() # Remove rows with NA values
+
+# change Date to Date object in R
+busData <- busData %>%
+  mutate(Date = as.Date(Date))
+
+# do the same as the above line but for subway data
+subwayData <- subwayData %>%
+  mutate(Date = as.Date(Date))
 
 # Create the directory for analysis data if it doesn't exist
 dir.create(here::here("./data/analysis_data"), showWarnings = FALSE)

### 2024-09-27 scripts/03-test_data.R

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Sanity check of the data
-# Author: Rohan Alexander
+# Author: Andrew Goh
 # Date: 19 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Contact: andrew.goh@utoronto.ca
 # License: MIT
 # Pre-requisites: Need to have simulated data
 # Any other information needed? None.
@@ -10,19 +10,109 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(testthat)
 
+# test simulated data
+data <- read_csv(here::here("data/raw_data/simulated.csv"))
 
-#### Test data ####
-data <- read_csv("data/raw_data/simulated.csv")
+# Convert Date column to Date format
+data$Date <- as.Date(data$Date)
 
-# Test for negative numbers
-data$number_of_marriage |> min() <= 0
+# Test 1: Check if all dates are within 2022-2023
+test_that("All dates are within 2022-2023", {
+  expect_true(all(data$Date >= as.Date("2022-01-01") & data$Date <= as.Date("2023-12-31")))
+})
 
-# Test for NAs
-all(is.na(data$number_of_marriage))
+# Test 2: Check if all Min.Delay values are non-negative
+test_that("No negative values in Min.Delay", {
+  expect_true(all(data$Min.Delay >= 0))
+})
 
+# Test 3: Check for any missing values in the data
+test_that("No missing values (NA) in the dataset", {
+  expect_false(any(is.na(data)))
+})
 
+# test raw data
+data <- read_csv(here::here("data/raw_data/raw_data.csv"))
 
+# Convert Date column to Date format
+data$Date <- as.Date(data$Date)
 
+# Test 1: Check if all dates are within 2022-2023
+test_that("All dates are within 2022-2023", {
+  expect_true(all(data$Date >= as.Date("2022-01-01") & data$Date <= as.Date("2023-12-31")))
+})
 
+# Test 2: Check if all Min.Delay values are non-negative
+test_that("No negative values in Min.Delay", {
+  expect_true(all(data$"Min Delay" >= 0))
+})
 
+# Test 3: Check for any missing values in the data
+test_that("No missing values (NA) in the dataset", {
+  expect_false(any(is.na(data)))
+})
+
+# test raw data for buses
+data <- read_csv(here::here("data/raw_data/raw_data_bus.csv"))
+
+# Convert Date column to Date format
+data$Date <- as.Date(data$Date)
+
[…]

### 2024-09-27 scripts/02-data_cleaning.R
updated script to clean out rows where min.delay was 0, then updated the data. also updated LLM usage to reflect the new code

@@ -10,6 +10,7 @@
 library(here)
 library(lubridate)
 library(dplyr)
+
 # Read the raw data
 subwayData <- read.csv(here::here("./data/raw_data/raw_data.csv"))
 busData <- read.csv(here::here("./data/raw_data/raw_data_bus.csv"))
@@ -22,20 +23,22 @@ busData$Time <- hm(busData$Time)
 subwayData$Hour <- hour(subwayData$Time)
 busData$Hour <- hour(busData$Time)
 
+# Select relevant columns and remove rows with NA values
 subwayData <- subwayData %>%
   select(Time, Date, Min.Delay, Hour) %>%
-  na.omit() # Remove rows with NA values
+  na.omit() %>%
+  filter(Min.Delay != 0) # Remove rows where Min.Delay is 0
 
 busData <- busData %>%
   select(Time, Date, Min.Delay, Hour) %>%
-  na.omit() # Remove rows with NA values
+  na.omit() %>%
+  filter(Min.Delay != 0) # Remove rows where Min.Delay is 0
 
-# change Date to Date object in R
-busData <- busData %>%
+# Convert Date column to Date object
+subwayData <- subwayData %>%
   mutate(Date = as.Date(Date))
 
-# do the same as the above line but for subway data
-subwayData <- subwayData %>%
+busData <- busData %>%
   mutate(Date = as.Date(Date))
 
 # Create the directory for analysis data if it doesn't exist

### 2024-09-27 paper/paper.qmd
Final paper and references.bib upload

--- 2024-09-23 paper/paper.qmd
@@
 ---
-title: "TTC delays: when they happen and how to be punctual"
+title: "Rush hour subways surprisingly excel at punctuality: a closer look at TTC Delays"
 subtitle: "Insight into when delays disproportionately tend to happen and related statistics"
@@
   - Andrew Goh
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: https://github.com/mushroomcarbon/TTCDelays."
 date: today
 date-format: long
-abstract: "TBA"
+abstract: "This paper uses data from Open Data Toronto on TTC subway and bus
+delays to analyze how the delays are generally distributed w.r.t. the time of day. Mean, total, and adjusted statistics were graphed, leading to the conclusions that subways are almost always
+more punctual than buses, peak hour leads to the fewest delays, and leaving 10-15 minutes earlier is a safe margin of error for delays.
+The conclusions of this paper provide a practical, actionable set of guidelines on how citizens can ensure/maximize punctuality when using the TTC to go about the city, as well as potential insight into where resources can be invested in to increase TTC efficiency the most."
 format: pdf
@@
 bibliography: references.bib
+header-includes:
+  - \usepackage{placeins}
 ---
 
+# Introduction
+
+The TTC, as Toronto's primary public transport network, is how 1.2 million people (@ttc_overview) get about the city every day.
+With its network of subways, the average TTC rider will be less affected by traffic than the average driver is, but unfortunately, delays act as public transport's traffic counterpart.
+And while individuals have their own ways to consistently remain punctual, there remains a lack of objective analysis regarding TTC delays.
+To fill this gap, we look into TTC data to determine just when and how delays happen, as well as how individuals can best make use of this information.
+
+Data on TTC delays for both subways and buses was obtained from opendatatoronto, then analysed and graphed.
+The delays were graphed both by mean delay time and total delay time (minutes) against the time of day (hour) to more intuitively see how the two explanatory variables, namely the type of transportation (subway/bus) and time of day (hour), affected various aspects of the delay experienced.
+Data was then adjusted for the difference in subway frequencies during peak vs.
+off hours and then graphed.
+
+From the graphs, we conclude that the mean delay times of buses is significantly higher than that of subways, whereas both transportation types have similar mean delays w.r.t.
+the time of day.
+Total delay times are higher during peak hours, but when accounting for the ~doubled amount of subways running during that time, peak hour subways actually end up being the least delayed ones on average.
+The significance of insight into public transportation delays lies both individually and socially: individuals can learn to better plan their trips to remain punctual in the event of delays, and considering how TTC delays can impact productivity as a whole on a large scale, governments and corporations alike can utilise this information to better manage resources to increase efficiency.
+
+The paper begins by introducing and viewing the dataset used, as well as analysing collection methodologies, larger contexts, as well as how the data was processed in Section 2.
+We then continue by graphing the data and discussing some preliminary insights regarding the visuals.
+In Section 3, we discuss the results of the data in a more in-depth fashion, as well as highlighting some key insights extrapolated from the previous analysis.
+The section then concludes by acknowledging the limitations of the study and discussing ways to further the research.
+
+# Data {#sec-mydatasection}
+
+The R programming language (@talia), dplyr (@dplyr), opendatatoronto(@opendatatoronto), and tidyverse(@thereferencecanbewhatever) were used to download, modify, and analyse data obtained from Open Data Toronto.
+Styler (@styler) was used to style the code.
+Data used in this paper comes from the "TTC Subway Delays" (@subway_data) and "TTC Bus Delays" (@bus_data) data sets from the Open Data Toronto database, published by the TTC.
+
+The two datasets are similar, providing the same key information: when delays happened (date and time), and how long they were delayed by.
+Cleaning the data involved removing N/A rows and rows where the delay was 0, then adding a new variable for the hour the delay happened in simply by extracting the hour from the more detailed time information provided.
+This sums up to 4 variables in the analysis dataset - Date, Time, and Hour describe when the delay happened, and Min Delay describes how long the delay was for, i.e.
+the difference between the time that the vehicle arrived and the time that the vehicle should have arrived.
+@tbl-analysis provides a short sample of the cleaned data, which is in the same format for both subway and bus datasets.
+Raw data for both datasets contained many more descriptors which were not used in this paper and is therefore shown in the Appendix in @tbl-raw-1 and @tbl-raw-2.
+
+<!-- the following table code written with assistance from Yisu Hou: https://github.com/YisuHou1/Toronto-Police-Budget-Statistics -->
+```{r, results='asis', echo=FALSE, fig.pos='H'}
+#| warning: false
+#| message: false
+#| echo: false
+# Load necessary libraries
+library(tidyverse)
+library(knitr)
+library(kableExtra)
+
+# Read the CSV file and select the first 8 rows
+data <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv")) %>%
+  slice(18:22)
+
+# Create a kable table with LaTeX format
+data %>%
+  kable("latex",
+    caption = "Sample of cleaned analysis data {#tbl-analysis}",
+    col.names = c("Time", "Date", "Min. Delay", "Hour"),
+    booktabs = TRUE
+  ) %>% # Use booktabs for better formatting
+  kable_styling(latex_options = c("striped", "hold_position")) # Add hold_position for placement
+```
+\FloatBarrier
+Data was taken from 2022-2023 due to the recency - data before 2021 was left out due to COVID-19 affecting statistics and data before COVID too dated for either to be relevant to 2024.
+There are no similar datasets that could have been used to analyze TTC delays - other TTC datasets don't describe the same thing and therefore can't be used either.
+
+The data was collected via the CIS (communications information system), the TTC's vehicle monitoring system (@methodology).
+This is done via GPS tracking on latitude-longitude data updated every 20 seconds, with erroneous data being autocorrected by the system.
+The TTC's method of measurement is reliable due to its high accuracy and error-tolerant nature, lending to the dataset's strong credibility.
+
+To have a broad understanding of the data, we first graph the summary statistics: @fig-times-mean shows the mean delay w.r.t. the hour, and @fig-times-stdev shows the standard deviations.
+
 ```{r}
-#| include: false
+#| label: fig-times-mean
+#| fig-cap: Subway delays w.r.t. time of day
+#| echo: false
 #| warning: false
@@
   group_by(Hour) %>%
-  summarise(Total_Delay_Subway = sum(Min.Delay, na.rm = TRUE), 
-            Mean_Delay_Subway = mean(Min.Delay, na.rm = TRUE))
+  summarise(
+    Total_Delay_Subway = sum(Min.Delay, na.rm = TRUE),
+    Mean_Delay_Subway = mean(Min.Delay, na.rm = TRUE)
+  )
 
@@
   group_by(Hour) %>%
-  summarise(Total_Delay_Bus = sum(Min.Delay, na.rm = TRUE), 
-            Mean_Delay_Bus = mean(Min.Delay, na.rm = TRUE))
+  summarise(
+    Total_Delay_Bus = sum(Min.Delay, na.rm = TRUE),
+    Mean_Delay_Bus = mean(Min.Delay, na.rm = TRUE)
+  )
 
@@
 hourly_avg_combined_long <- hourly_delays_combined %>%
-  pivot_longer(cols = starts_with("Mean"), 
+  pivot_longer(
+    cols = starts_with("Mean"),
     names_to = "Type",
-               values_to = "Mean_Delay") %>%
+    values_to = "Mean_Delay"
+  ) %>%
   mutate(Type = recode(Type,
     "Mean_Delay_Subway" = "Subway",
-                       "Mean_Delay_Bus" = "Bus"))
+    "Mean_Delay_Bus" = "Bus"
+  ))
 
@@
   geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Average Delays by Hour of Day (Subway vs Bus)", 
+  labs(
+    title = "Average Delays by Hour of Day (Subway vs Bus)",
     x = "Hour of Day",
     y = "Average Delay (Minutes)",
-       fill = "Type") +
+    fill = ""
+  ) +
   scale_fill_manual(values = c("Subway" = "steelblue", "Bus" = "tomato")) +
   theme_minimal() +
-  theme(legend.position = "top", 
-        plot.title = element_text(hjust = 0.5, face = "bold"))
+  theme(
+    legend.position = "top",
+    plot.title = element_text(hjust = 0.5, face = "bold")
+  )
 ```
+
+```{r}
+#| label: fig-times-stdev
+#| fig-cap: Standard deviation of subway and bus delays w.r.t. time of day
+#| echo: false
+#| warning: false
+#| message: false
+# Calculate standard deviation for subway and bus data by hour
+hourly_sd_subway <- subwayData %>%
+  group_by(Hour) %>%
+  summarise(
+    StdDev_Delay_Subway = sd(Min.Delay, na.rm = TRUE)
+  )
+
+hourly_sd_bus <- busData %>%
+  group_by(Hour) %>%
+  summarise(
+    StdDev_Delay_Bus = sd(Min.Delay, na.rm = TRUE)
+  )
+
+# Combine subway and bus data for standard deviation comparison
+hourly_sd_combined <- hourly_sd_subway %>%
+  full_join(hourly_sd_bus, by = "Hour")
+
+# Reshape data for standard deviations
+hourly_sd_combined_long <- hourly_sd_combined %>%
+  pivot_longer(
+    cols = starts_with("StdDev"),
+    names_to = "Type",
+    values_to = "StdDev_Delay"
+  ) %>%
+  mutate(Type = recode(Type,
+    "StdDev_Delay_Subway" = "Subway",
+    "StdDev_Delay_Bus" = "Bus"
+  ))
+
+# Plot 2: Standard deviation of delays by hour of day (combined subway and bus)
+ggplot(hourly_sd_combined_long, aes(x = factor(Hour), y = StdDev_Delay, fill = Type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(
+    title = "Standard Deviation of Delays by Hour of Day (Subway vs Bus)",
+    x = "Hour of Day",
+    y = "Standard Deviation of Delay (Minutes)",
+    fill = ""
+  ) +
+  scale_fill_manual(values = c("Subway" = "steelblue", "Bus" = "tomato")) +
+  theme_minimal() +
+  theme(
+    legend.position = "top",
+    plot.title = element_text(hjust = 0.5, face = "bold")
+  )
+```
+\FloatBarrier
+From the mean graph (@fig-times-mean) we can see that 1.
+the average magnitude of delays is generally homogeneous within transportation categories, and that 2.
+the average delay times for buses is around 4-5 times more than that of subways.
+This confirms the intuition that subways are generally more time-reliable than buses, and we will therefore focus on subways from hereonafter to find the ideal, delay-minimising way of transport.
+The standard deviation graph (@fig-times-stdev) shows that the standard deviation of both bus and subway delays are all larger than the means of the respective measurements.
+This suggests that the mean itself, while still an accurate estimate of typical delays, holds less practical significance as a benchmark for normalisation (i.e.
+leaving x minutes early to offset delay).
+
+While there is no data regarding the total number of subway trips and therefore cannot calculate an "average delay per ride" statistic, it is still possible to sum up the delay at each hour of day to obtain an estimator for the relative amount of delay encountered during each respective hour, as shown in @fig-times-total.
+
 ```{r}
 #| label: fig-times-total
-#| fig-cap: total delay times
+#| fig-cap: Total delay times of the TTC subway system per hour
 #| echo: false
+
+library(tidyverse)
+library(dplyr)
+library(lubridate)
+library(here)
+
+# Load the cleaned data
+subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
+busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
 
@@
   geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Total Subway Delays by Hour of Day", 
+  labs(
+    title = "Total Subway Delays by Hour of Day",
     x = "Hour of Day",
-       y = "Total Delay (Minutes)") +
+    y = "Total Delay (Minutes)"
+  ) +
   theme_minimal() +
@@
 ```
+\FloatBarrier
+The data in @fig-times-total singles out 6-9am and 4-6pm to be the worst offenders, suggesting the conclusion that peak hour riders are the most susceptible to TTC-induced tardiness.
+According to the @ttc_line_1_schedule, though, subways run at approximately double the frequency during these rush hours, accounting for the increase in total delay times by sheer volume.
+Adjusting for this by halving the relative delay times for the aforementioned rush hours (6-9am, 3-7pm), we obtain @fig-times-adjusted, the adjusted total delay graph.
+
 ```{r}
-#| label: fig-times-adjustedd
-#| fig-cap: delay time adjusted for the approximate subway load
+#| label: fig-times-adjusted
+#| fig-cap: Delay time adjusted for the approximate subway load
 #| echo: false
+
+library(tidyverse)
+library(dplyr)
+library(lubridate)
+library(here)
+
+# Load the cleaned data
+subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
+busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
+
+# Function to adjust subway delays
 adjust_subway_delays <- function(data) {
@@
   geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Total Subway Delays by Hour of Day Adjusted for Load Differences", 
+  labs(
+    title = "Total Subway Delays by Hour, Adjusted for Capacity",
     x = "Hour of Day",
-       y = "Delay (Minutes)") +
+    y = "Delay (Minutes)"
+  ) +
   theme_minimal() +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))
-
 ```
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
+\FloatBarrier
 # Discussion
 
-## First discussion point {#sec-first-point}
+Looking at @fig-times-adjusted, peak hours are, surprisingly, the least affected by delays on average.
+Combining this with previous conclusions, we get that peak-hour subways are the most time-reliable vehicles of the TTC.
+A potential reason for this could be the higher-stress situation leading to tighter schedules, the lack of shift changes, or even the general alertness of drivers, among others - this, however, is not the focus nor concern of our current investigation.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+The simple conclusion of this is that students and workers regularly making use of the peak hour subways have no need to change their TTC usage - individuals making more flexible plans, however, are incentivised to head out during peak hours instead of avoiding them, if speed is the key concern.
 
-## Second discussion point
+A key implication of this that ties into the broader context is that with the limited resources of the TTC (@ttc_conventional_system_2023), focusing more resources into peak hours rather than spreading them out evenly is the most economic decision, barring other considerations: peak hour subways manage the highest efficiency of all lines currently, whether it be with respect to passenger load or time to arrival; therefore, increasing the frequency of peak-hour subways would do the most to reduce passengers' times in transit and increase the city's functionality as a whole.
 
-## Third discussion point
+A key question that remains, however, is how individuals can best use this data to inform their own travels.
+As mentioned in Section 2, a high standard deviation suggests that a simple mean offset is insufficient/nonideal to offset the effect of delays.
+Considering how delays are measured, however, notice how a single subway would rack up delays as it operates - if its arrival at station 1 is delayed by 5 minutes, for example, its successive arrivals to future stations will all also reflect this "carried-over" delay.
+Subway delays, therefore, follow a roughly monotononically increasing pattern, which would largely explain the disproportionately large standard deviation.
 
-## Weaknesses and next steps
+To better estimate the average delay amount given that a delay happens, we look at average ridership data from @metrolinx_rtp and @average_distance, stating that "local" trips take around 4.2km on average.
+Average distance between stations, calculated using data from the @ttc_conventional_system_2023, is 70 stations / 70.1km for approximately 1km per station (on average) and a resulting average of around 4 stations per trip.
 
-Weaknesses and next steps should also be included.
+Accounting for the previously mentioned carry-over delay for the specific subway arriving at each of the 4 stations, we conclude that for the average trip during peak hour, leaving 10-15 minutes early is a fairly safe margin to ensure punctual arrival even in the case of delays, doubling this value to account for extra stations (20-30 minutes extra for an 8-station trip, for example).
+
+# Limitations and further research
+
+A limitation of this paper is how, as mentioned in Section 2, there is a lack of information on the absolute (as opposed to relative) frequency of subway trips.
+This means that average-case scenarios can only be calculated in the conditional case (i.e.
+"given that a delay happens"), rather than factoring for the total probability of delays happening.
+A further limitation is the lack of the most recent data - data from 2024 was omitted to avoid introducing bias to the model due to the year, as the data from 2024 is missing data from the final three months for obvious reasons.
+A repeat of the study, conducted on the complete 2024 dataset after september, could better reflect recent trends and patterns.
+
+Future studies could focus on the omitted columns from the raw data, analysing how factors such as the station where delays happen, model of bus, and direction of travel affect the magnitude of delays.
+Adding more variables to the analysis would likely improve the insights obtained from the data - fitting a statistical model (ANOVA, KMeans, etc.) could also be a potential route to take to obtain some more precise, methodical conclusions.
 
 \newpage
+# Appendix
 
-\appendix
+```{r, echo=FALSE, fig.pos='H'}
+#| warning: false
+#| message: false
+#| echo: false
+# Load necessary libraries
+library(tidyverse)
+library(knitr)
+library(kableExtra)
 
-# Appendix {-}
+# Read the CSV file and select the first 8 rows
+data <- read.csv(here::here("./data/raw_data/raw_data.csv")) %>%
+  slice(1:8)
 
-# Additional data details
+# Split the data into two parts
+data1 <- data %>% select(Date, Time, Day, Station, Code) # First half of the columns
+data2 <- data %>% select(`Min.Delay`, `Min.Gap`, Bound, Line, Vehicle) # Second half of the columns
+
+# Create the first table
+table1 <- data1 %>%
+  kable("latex",
+    caption = "Sample of raw data {#tbl-raw-1}",
+    col.names = c("Date", "Time", "Day", "Station", "Code"),
+    booktabs = TRUE
+  ) %>%
+  kable_styling(latex_options = c("striped", "hold_position"))
+table1
+```
+
+```{r, echo=FALSE, fig.pos='H'}
+#| warning: false
+#| message: false
+#| echo: false
+#|
+table2 <- data2 %>%
+  kable("latex",
+    caption = "Sample of raw data (continued) {#tbl-raw-2}",
+    col.names = c("Min Delay", "Min Gap", "Bound", "Line", "Vehicle"),
+    booktabs = TRUE
+  ) %>%
+  kable_styling(latex_options = c("striped", "hold_position"))
+table2
+```
 
 \newpage
-
 # References

### 2024-09-27 paper/paper.qmd
final, minor, modifications to the paper

@@
 
-Accounting for the previously mentioned carry-over delay for the specific subway arriving at each of the 4 stations, we conclude that for the average trip during peak hour, leaving 10-15 minutes early is a fairly safe margin to ensure punctual arrival even in the case of delays, doubling this value to account for extra stations (20-30 minutes extra for an 8-station trip, for example).
+Accounting for the previously mentioned carry-over delay for the specific subway arriving at each of the 4 stations, we conclude that for the average trip during peak hour, leaving 10-15 minutes early is a fairly safe margin to ensure punctual arrival even in the case of delays, scaling this value as fit to account for extra/fewer stations (leave a 20-30 minutes margin for an 8-station trip, for example).
 

### 2024-10-10 scripts/01-download_data.R
all

@@ -1,47 +1,14 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from Open Data Toronto
+# Purpose: Downloads and saves the data from Spotify
 # Author: Andrew Goh
-# Date: 27 September 2024
+# Date: 11 October 2024
 # Contact: andrew.goh@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
 # Any other information needed? None
 
-library(opendatatoronto)
 library(dplyr)
 library(tidyverse)
 
-#---------------subway data:----------------#
-
-# get package
-package <- show_package("996cfe8d-fb35-40ce-b569-698d51fc683b")
-
-# get all resources for this package
-resources <- list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c("csv", "geojson"))
-data <- filter(resources, row_number() == 9) %>%
-  get_resource()
-data2 <- filter(resources, row_number() == 10) %>%
-  get_resource()
-# merge 2022 and 2023 data into one set
-data <- rbind(data, data2)
-# save into csv
-write_csv(data, "data/raw_data/raw_data.csv")
-
-#---------------bus data:----------------#
-
-# do the same for bus data: get data
-package <- show_package("e271cdae-8788-4980-96ce-6a5c95bc6618")
-# identify resources
-resources <- list_package_resources("e271cdae-8788-4980-96ce-6a5c95bc6618")
-# get data
-data <- filter(resources, row_number() == 10) %>%
-  get_resource()
-data2 <- filter(resources, row_number() == 11) %>%
-  get_resource()
-# merge
-data <- rbind(data, data2)
-# save
-write_csv(data, "data/raw_data/raw_data_bus.csv")
+the_national <- get_artist_audio_features("the national")
+saveRDS(the_national, "data/raw_data/the_national.rds")

### 2024-10-10 scripts/01-download_data.R
edited date

@@ -1,14 +1,16 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Spotify
 # Author: Andrew Goh
-# Date: 11 October 2024
+# Date: 10 October 2024
 # Contact: andrew.goh@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
 # Any other information needed? None
-
+library(spotifyr)
 library(dplyr)
 library(tidyverse)
 
 the_national <- get_artist_audio_features("the national")
-saveRDS(the_national, "data/raw_data/the_national.rds")
+saveRDS(the_national, "data/the_national.rds")
+
+

### 2024-10-10 scripts/02-data_cleaning.R
edited date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Cleans the raw delay data into an analysis dataset
 # Author: Andrew Goh
-# Date: 19 September 2024
+# Date: 10 October 2024
 # Contact: andrew.goh@utoronto.ca
 # License: MIT
 # Pre-requisites: Need to have downloaded the data

### 2024-10-10 scripts/03-test_data.R
added relevant tests on Spotify dataset

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Sanity check of the data
 # Author: Andrew Goh
-# Date: 19 September 2024
+# Date: 10 October 2024
 # Contact: andrew.goh@utoronto.ca
 # License: MIT
 # Pre-requisites: Need to have simulated data
@@ -10,109 +10,17 @@
 
 #### Workspace setup ####
 library(tidyverse)
-library(testthat)
 
-# test simulated data
-data <- read_csv(here::here("data/raw_data/simulated.csv"))
+# test data from Spotify
+data <- read_rds(here::here("data/the_national.rds"))
 
-# Convert Date column to Date format
-data$Date <- as.Date(data$Date)
+# test for missing values in the dataset
+any(is.na(data))
 
-# Test 1: Check if all dates are within 2022-2023
-test_that("All dates are within 2022-2023", {
-  expect_true(all(data$Date >= as.Date("2022-01-01") & data$Date <= as.Date("2023-12-31")))
-})
+# test whether variables of interest are doubles
+is.double(data$album_release_year)
+is.double(data$danceability)
+is.double(data$energy)
+is.double(data$instrumentalness)
+is.double(data$valence)
 
-# Test 2: Check if all Min.Delay values are non-negative
-test_that("No negative values in Min.Delay", {
-  expect_true(all(data$Min.Delay >= 0))
-})
-
-# Test 3: Check for any missing values in the data
-test_that("No missing values (NA) in the dataset", {
-  expect_false(any(is.na(data)))
-})
-
-# test raw data
-data <- read_csv(here::here("data/raw_data/raw_data.csv"))
-
-# Convert Date column to Date format
-data$Date <- as.Date(data$Date)
-
-# Test 1: Check if all dates are within 2022-2023
-test_that("All dates are within 2022-2023", {
-  expect_true(all(data$Date >= as.Date("2022-01-01") & data$Date <= as.Date("2023-12-31")))
-})
-
-# Test 2: Check if all Min.Delay values are non-negative
-test_that("No negative values in Min.Delay", {
-  expect_true(all(data$"Min Delay" >= 0))
-})
-
-# Test 3: Check for any missing values in the data
-test_that("No missing values (NA) in the dataset", {
-  expect_false(any(is.na(data)))
-})
-
-# test raw data for buses
-data <- read_csv(here::here("data/raw_data/raw_data_bus.csv"))
-
-# Convert Date column to Date format
[…]

### 2024-10-10 paper/paper.qmd
updated plots

--- 2024-09-23 paper/paper.qmd
@@
 ---
-title: "TTC delays: when they happen and how to be punctual"
+title: "Rush hour subways surprisingly excel at punctuality: a closer look at TTC Delays"
 subtitle: "Insight into when delays disproportionately tend to happen and related statistics"
 author: 
-  - Andrew Goh
-thanks: "Code and data are available at: LINK. @talia, @thereferencecanbewhatever, @dplyr were used in this paper."
+  - Andrew Goh, Yisu Hou, Jiwon Choi
+thanks: "Code and data are available at: https://github.com/mushroomcarbon/TTCDelays."
 date: today
 date-format: long
-abstract: "This paper uses data from Open Data Toronto (@opendatatoronto) on TTC subway and bus
-delays to analyze how the delays are generally distributed w.r.t. the time of day. Mean, total, and adjusted
-statistics are graphed and discussed to reach a conclusion."
+abstract: "This paper uses data from Open Data Toronto on TTC subway and bus
+delays to analyze how the delays are generally distributed w.r.t. the time of day. Mean, total, and adjusted statistics were graphed, leading to the conclusions that subways are almost always
+more punctual than buses, peak hour leads to the fewest delays, and leaving 10-15 minutes earlier is a safe margin of error for delays.
+The conclusions of this paper provide a practical, actionable set of guidelines on how citizens can ensure/maximize punctuality when using the TTC to go about the city, as well as potential insight into where resources can be invested in to increase TTC efficiency the most."
 format: pdf
@@
 bibliography: references.bib
+header-includes:
+  - \usepackage{placeins}
 ---
@@
 
-The TTC, as Toronto's primary public transport network, is the means for many
-to get about in the city. At an annual cost of ~$1.4k CAD, it remains the vastly more 
-economical means of transport in Toronto, especially compared to driving costs
-averaging $13k-16k per year, according to a CAA estimate by the @ttc_know_your_ttc.
-
-With its network of subways, the average TTC rider will be, comparatively, less affected by traffic as
-the average driver is, but all is not always smooth sailing with the TTC either. Delays, whether
-it be in the bus or subway systems, are decently common in the TTC family of transportation devices,
-with each delay potentially ruining an hour's worth of brisk preparation and a day's worth of planning.
-
-While everyone has their own understanding of delays, with discrepancies amplified by personal
-experience and survivor's bias, we look into the data to determine just when and how delays happen.
+Write Introduction here
 
@@
 
-Armed with the simple driving question of how I can reduce delays in my day-to-day transits,
-we first graph the mean subway and bus delay data (@fig-times-mean) to have a broad understanding of the data.
+The R programming language (@talia), dplyr (@dplyr), opendatatoronto(@opendatatoronto), and tidyverse(@thereferencecanbewhatever) were used to download, modify, and analyse data obtained from Open Data Toronto.
+Styler (@styler) was used to style the code.
+Data used in this paper comes from the "TTC Subway Delays" (@subway_data) and "TTC Bus Delays" (@bus_data) data sets from the Open Data Toronto database, published by the TTC.
+
 ```{r}
-#| label: fig-times-mean
-#| fig-cap: Subway delays w.r.t. time of day
+#| label: fig-dance
+#| fig-cap: Danceability Over Album Release Years
 #| echo: false
@@
 library(dplyr)
-library(lubridate)
 library(here)
 
-# Load the cleaned data
-subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
-busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
+# Load the data
+data <- read_rds(here::here("data/the_national.rds"))
 
-# Aggregate data by hour to calculate total and mean delays
-hourly_delays_subway <- subwayData %>%
-  group_by(Hour) %>%
-  summarise(Total_Delay_Subway = sum(Min.Delay, na.rm = TRUE), 
-            Mean_Delay_Subway = mean(Min.Delay, na.rm = TRUE))
-
-hourly_delays_bus <- busData %>%
-  group_by(Hour) %>%
-  summarise(Total_Delay_Bus = sum(Min.Delay, na.rm = TRUE), 
-            Mean_Delay_Bus = mean(Min.Delay, na.rm = TRUE))
-
-# Combine subway and bus data for comparison
-hourly_delays_combined <- hourly_delays_subway %>%
-  full_join(hourly_delays_bus, by = "Hour")
-
-# Reshape data for average delays
-hourly_avg_combined_long <- hourly_delays_combined %>%
-  pivot_longer(cols = starts_with("Mean"), 
-               names_to = "Type", 
-               values_to = "Mean_Delay") %>%
-  mutate(Type = recode(Type, 
-                       "Mean_Delay_Subway" = "Subway", 
-                       "Mean_Delay_Bus" = "Bus"))
-
-# Plot 1: Average delays by hour of day (combined subway and bus)
-ggplot(hourly_avg_combined_long, aes(x = factor(Hour), y = Mean_Delay, fill = Type)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Average Delays by Hour of Day (Subway vs Bus)", 
-       x = "Hour of Day", 
-       y = "Average Delay (Minutes)", 
-       fill = "Type") +
-  scale_fill_manual(values = c("Subway" = "steelblue", "Bus" = "tomato")) +
-  theme_minimal() +
-  theme(legend.position = "top", 
-        plot.title = element_text(hjust = 0.5, face = "bold"))
+# Plot 1: danceability over time
+ggplot(data, aes(x = album_release_year, y = danceability)) +
+  geom_point(alpha = 0.7, color = "black") + 
+  stat_summary(fun = mean, geom = "line", color = "green", size = 1.2) + 
+  labs(
+    x = "Album Release Year",
+    y = "Danceability"
+  ) +
+  theme_minimal()
 ```
-Two main takeaways from the figure: one, that the average magnitude of delays is generally the same over the day, whether it be for subways or buses, and two, that the average delay time for buses is around 4-5 times more than that of subways.
-This confirms the intuition that subways generally are a more time-reliable way to get somewhere than buses, and we will therefore focus on subways from hereonwards in our quest for the most reliable transit plan.
-
-While we don't know the total frequency of subways and therefore cannot calculate an "average delay per ride" statistic, what we can do is sum up the delay at each hour of day to still obtain an estimator for how much delay you're likely to encounter in a trip during each respective hour, as shown in @fig-times-total.
-```{r}
-#| label: fig-times-total
-#| fig-cap: total delay times
-#| echo: false
-
-library(tidyverse)
-library(dplyr)
-library(lubridate)
-library(here)
-
-# Load the cleaned data
-subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
-busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
-
-# Reshape data for total delays
-hourly_delays_subway <- subwayData %>%
-  group_by(Hour) %>%
-  summarise(Total_Delay_Subway = sum(Min.Delay, na.rm = TRUE))
-
-ggplot(hourly_delays_subway, aes(x = factor(Hour), y = Total_Delay_Subway)) +
-  geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Total Subway Delays by Hour of Day", 
-       x = "Hour of Day", 
-       y = "Total Delay (Minutes)") +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
-```
-From the data here, it seems that 8-9am and 3-5pm seem to be the worst offenders, which would suggest that
-transit during these times is the most prone to delays.
-According to the @ttc_line_1_schedule, though, subways run at approximately double the frequency during these rush hours, accounting for the increase in total delay times by sheer volume.
-Finally, adjusting for this by halving the relative delay times for the aforementioned rush hours (6-9am, 3-7pm), we obtain @fig-times-adjusted.
 
 ```{r}
-#| label: fig-times-adjusted
-#| fig-cap: Delay time adjusted for the approximate subway load
+#| label: fig-energy
+#| fig-cap: Energy Over Album Release Years
 #| echo: false
+#| warning: false
+#| message: false
 
-library(tidyverse)
-library(dplyr)
-library(lubridate)
-library(here)
-
-# Load the cleaned data
-subwayData <- read.csv(here::here("./data/analysis_data/analysis_data.csv"))
-busData <- read.csv(here::here("./data/analysis_data/analysis_data_bus.csv"))
-
-# Function to adjust subway delays
-adjust_subway_delays <- function(data) {
-  data %>%
-    mutate(Adjusted_Delay = ifelse(Hour %in% 6:9 | Hour %in% 15:19, Min.Delay / 2, Min.Delay))
-}
-
-# Apply the adjustment to subway data
-subwayData <- adjust_subway_delays(subwayData)
-
-# Aggregate the adjusted data by hour for total delays
-hourly_delays_subway <- subwayData %>%
-  group_by(Hour) %>%
-  summarise(Total_Delay_Subway = sum(Adjusted_Delay, na.rm = TRUE))
-
-# Plot: Adjusted total subway delays by hour of day
-ggplot(hourly_delays_subway, aes(x = factor(Hour), y = Total_Delay_Subway)) +
-  geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Total Subway Delays by Hour of Day Adjusted for Capacity", 
-       x = "Hour of Day", 
-       y = "Delay (Minutes)") +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
-
+# Plot 2: energy over time
+ggplot(data, aes(x = album_release_year, y = energy)) +
+  geom_point(alpha = 0.7, color = "black") +
+  stat_summary(fun = mean, geom = "line", color = "blue", size = 1.2) +
+  labs(
+    x = "Album Release Year",
+    y = "Energy"
+  ) +
+  theme_minimal()
 ```
 
-# Discussion
+```{r}
+#| label: fig-instrumental
+#| fig-cap: Instrumentalness Over Album Release Years
+#| echo: false
+#| warning: false
+#| message: false
 
-Surprisingly, peak hours are the most consistent hours of them all, at least when compared to the off-hours of the TTC.
-A potential reason for this could be the higher-stress situation leading to tighter schedules, the lack of shift changes, or even the general alertness of drivers, among others - this, however, is not the focus nor concern of our current investigation.
+# Plot 3: instrumentalness over time
+ggplot(data, aes(x = album_release_year, y = instrumentalness)) +
+  geom_point(alpha = 0.7, color = "black") +
+  stat_summary(fun = mean, geom = "line", color = "red", size = 1.2) +
+  labs(
+    x = "Album Release Year",
+    y = "Instrumentalness"
+  ) +
+  theme_minimal()
+```
 
-As to the average delay amount, given that a delay happens, we look at average ridership data from @metrolinx_rtp and @average_distance, stating that "local" trips take around 4.2km on average.
-Average distance between stations, calculated using data from @ttc_conventional_system_2023, is 70 stations / 70.1km for approximately 1km per station (on average) and a resulting average of around 4 stations per trip.
+```{r}
+#| label: fig-valence
+#| fig-cap: Valence Over Album Release Years
+#| echo: false
+#| warning: false
+#| message: false
 
-Assuming a carry-over delay for your specific subway arriving at each of the 4 stations, we conclude that for the average trip during peak hour, leaving 10-15 minutes early is a fairly safe margin to ensure punctual arrival even in the case of delays.
-\newpage
-
-# References
+# Plot 4: valence over time
+ggplot(data, aes(x = album_release_year, y = valence)) +
+  geom_point(alpha = 0.7, color = "black") +
+  stat_summary(fun = mean, geom = "line", color = "orange", size = 1.2) +
+  labs(
+    x = "Album Release Year",
+    y = "Valence"
+  ) +
+  theme_minimal()
+```
+\FloatBarrier

### 2024-10-10 scripts/03-test_data.R
added test code for plots (need to delete)

@@ -24,3 +24,50 @@ is.double(data$energy)
 is.double(data$instrumentalness)
 is.double(data$valence)
 
+
+
+# plot of danceability over time
+ggplot(data, aes(x = album_release_year, y = danceability)) +
+  geom_point(alpha = 0.7, color = "black") + 
+  stat_summary(fun = mean, geom = "line", color = "green", size = 1.2) + 
+  labs(
+    title = "Danceability Over Album Release Years",
+    x = "Album Release Year",
+    y = "Danceability"
+  ) +
+  theme_minimal()
+
+# plot of energy over time
+ggplot(data, aes(x = album_release_year, y = energy)) +
+  geom_point(alpha = 0.7, color = "black") +
+  stat_summary(fun = mean, geom = "line", color = "blue", size = 1.2) +
+  labs(
+    title = "Energy Over Album Release Years",
+    x = "Album Release Year",
+    y = "Energy"
+  ) +
+  theme_minimal()
+
+# plot of instrumentalness over time
+ggplot(data, aes(x = album_release_year, y = instrumentalness)) +
+  geom_point(alpha = 0.7, color = "black") +
+  stat_summary(fun = mean, geom = "line", color = "red", size = 1.2) +
+  labs(
+    title = "Instrumentalness Over Album Release Years",
+    x = "Album Release Year",
+    y = "Instrumentalness"
+  ) +
+  theme_minimal()
+
+# plot of valence over time
+ggplot(data, aes(x = album_release_year, y = valence)) +
+  geom_point(alpha = 0.7, color = "black") +
+  stat_summary(fun = mean, geom = "line", color = "orange", size = 1.2) +
+  labs(
+    title = "Valence Over Album Release Years",
+    x = "Album Release Year",
+    y = "Valence"
+  ) +
+  theme_minimal()
+
+

### 2024-10-10 scripts/00-simulate_data.R
updated preamble

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates data
-# Author: Andrew Goh
-# Date: 19 September 2024
+# Author: Andrew Goh, Yisu Hou, Jiwon Choi
+# Date: 10 October 2024
 # Contact: andrew.goh@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None

### 2024-10-10 scripts/01-download_data.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Spotify
-# Author: Andrew Goh
+# Author: Andrew Goh, Yisu Hou, Jiwon Choi
 # Date: 10 October 2024
 # Contact: andrew.goh@mail.utoronto.ca
 # License: MIT

### 2024-10-10 scripts/02-data_cleaning.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Cleans the raw delay data into an analysis dataset
-# Author: Andrew Goh
+# Author: Andrew Goh, Yisu Hou, Jiwon Choi
 # Date: 10 October 2024
 # Contact: andrew.goh@utoronto.ca
 # License: MIT

### 2024-10-10 scripts/03-test_data.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Sanity check of the data
-# Author: Andrew Goh
+# Author: Andrew Goh, Yisu Hou, Jiwon Choi
 # Date: 10 October 2024
 # Contact: andrew.goh@utoronto.ca
 # License: MIT

### 2024-10-10 scripts/00-simulate_data.R
modified scripts to clean & test the data

@@ -1,32 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates data
-# Author: Andrew Goh, Yisu Hou, Jiwon Choi
-# Date: 10 October 2024
-# Contact: andrew.goh@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: None
-# Any other information needed? None
-
-
-#### Workspace setup ####
-library(tidyverse)
-
-
-#### Simulate data ####
-set.seed(6737151)
-
-n <- 100
-# simulate 100 dates
-dates <- seq.Date(from = as.Date("2022-01-01"), by = "day", length.out = n)
-# same for times
-times <- format(
-  as.POSIXct("2022-01-01 00:00:00") + runif(n, min = 0, max = 24 * 3600),
-  format = "%H:%M:%S"
-)
-# delay in minutes
-min_delay <- abs(round(rnorm(n, 15, 5), 0))
-# combine
-simulated_data <- data.frame(Date = dates, Time = times, Min.Delay = min_delay)
-
-# write to file
-write_csv(simulated_data, file = "data/raw_data/simulated.csv")

### 2024-10-10 scripts/01-download_data.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Spotify
-# Author: Andrew Goh, Yisu Hou, Jiwon Choi
+# Author: Andrew Goh, Yisu Hou, Jiwon Choi, Xiaolu Ji
 # Date: 10 October 2024
 # Contact: andrew.goh@mail.utoronto.ca
 # License: MIT

### 2024-10-10 scripts/02-data_cleaning.R

@@ -1,6 +1,6 @@
 #### Preamble ####
-# Purpose: Cleans the raw delay data into an analysis dataset
-# Author: Andrew Goh, Yisu Hou, Jiwon Choi
+# Purpose: Cleans the raw spotify data into an analysis dataset
+# Author: Andrew Goh, Yisu Hou, Jiwon Choi, Xiaolu Ji
 # Date: 10 October 2024
 # Contact: andrew.goh@utoronto.ca
 # License: MIT
@@ -11,39 +11,16 @@ library(here)
 library(lubridate)
 library(dplyr)
 
-# Read the raw data
-subwayData <- read.csv(here::here("./data/raw_data/raw_data.csv"))
-busData <- read.csv(here::here("./data/raw_data/raw_data_bus.csv"))
+# Load the data
+data <- read_rds(here::here("data/raw_data/the_national.rds"))
 
-# Convert the Time column to a proper time format
-subwayData$Time <- hm(subwayData$Time) # Parse 'Time' as hours and minutes
-busData$Time <- hm(busData$Time)
+# Ensure that variables of interest are doubles, removing rows if not
+data <- data %>%
+  filter(is.double(album_release_year),
+         is.double(danceability),
+         is.double(energy),
+         is.double(instrumentalness),
+         is.double(valence))
 
-# Extract the hour from the Time column
-subwayData$Hour <- hour(subwayData$Time)
-busData$Hour <- hour(busData$Time)
-
-# Select relevant columns and remove rows with NA values
-subwayData <- subwayData %>%
-  select(Time, Date, Min.Delay, Hour) %>%
-  na.omit() %>%
-  filter(Min.Delay != 0) # Remove rows where Min.Delay is 0
-
-busData <- busData %>%
-  select(Time, Date, Min.Delay, Hour) %>%
-  na.omit() %>%
-  filter(Min.Delay != 0) # Remove rows where Min.Delay is 0
-
-# Convert Date column to Date object
-subwayData <- subwayData %>%
-  mutate(Date = as.Date(Date))
-
-busData <- busData %>%
-  mutate(Date = as.Date(Date))
-
-# Create the directory for analysis data if it doesn't exist
-dir.create(here::here("./data/analysis_data"), showWarnings = FALSE)
-
-# Write the cleaned data to new CSV files
-write.csv(subwayData, here::here("./data/analysis_data/analysis_data.csv"), row.names = FALSE)
[…]

### 2024-10-10 scripts/03-test_data.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Sanity check of the data
-# Author: Andrew Goh, Yisu Hou, Jiwon Choi
+# Author: Andrew Goh, Yisu Hou, Jiwon Choi, Xiaolu Ji
 # Date: 10 October 2024
 # Contact: andrew.goh@utoronto.ca
 # License: MIT
@@ -12,7 +12,7 @@
 library(tidyverse)
 
 # test data from Spotify
-data <- read_rds(here::here("data/the_national.rds"))
+data <- read_rds(here::here("data/analysis_data/the_national_cleaned.rds"))
 
 # test for missing values in the dataset
 any(is.na(data))
@@ -23,51 +23,4 @@ is.double(data$danceability)
 is.double(data$energy)
 is.double(data$instrumentalness)
 is.double(data$valence)
-
-
-
-# plot of danceability over time
-ggplot(data, aes(x = album_release_year, y = danceability)) +
-  geom_point(alpha = 0.7, color = "black") + 
-  stat_summary(fun = mean, geom = "line", color = "green", size = 1.2) + 
-  labs(
-    title = "Danceability Over Album Release Years",
-    x = "Album Release Year",
-    y = "Danceability"
-  ) +
-  theme_minimal()
-
-# plot of energy over time
-ggplot(data, aes(x = album_release_year, y = energy)) +
-  geom_point(alpha = 0.7, color = "black") +
-  stat_summary(fun = mean, geom = "line", color = "blue", size = 1.2) +
-  labs(
-    title = "Energy Over Album Release Years",
-    x = "Album Release Year",
-    y = "Energy"
-  ) +
-  theme_minimal()
-
-# plot of instrumentalness over time
-ggplot(data, aes(x = album_release_year, y = instrumentalness)) +
-  geom_point(alpha = 0.7, color = "black") +
-  stat_summary(fun = mean, geom = "line", color = "red", size = 1.2) +
-  labs(
-    title = "Instrumentalness Over Album Release Years",
-    x = "Album Release Year",
-    y = "Instrumentalness"
-  ) +
-  theme_minimal()
-
-# plot of valence over time
-ggplot(data, aes(x = album_release_year, y = valence)) +
-  geom_point(alpha = 0.7, color = "black") +
-  stat_summary(fun = mean, geom = "line", color = "orange", size = 1.2) +
-  labs(
-    title = "Valence Over Album Release Years",
[…]

### 2024-10-10 paper/paper.qmd
styler

@@
 ---
-title: "Rush hour subways surprisingly excel at punctuality: a closer look at TTC Delays"
-subtitle: "Insight into when delays disproportionately tend to happen and related statistics"
+title: "Analyzing The National's Musical Trajectory from 2001 to 2023"
+subtitle: "A Spotify Data Analysis of Danceability, Energy, Instrumentalness, and Valence Trends"
 author: 
-  - Andrew Goh, Yisu Hou, Jiwon Choi
-thanks: "Code and data are available at: https://github.com/mushroomcarbon/TTCDelays."
+  - Andrew Goh, Yisu Hou, Jiwon Choi, Xiaolu Ji
+thanks: "Code and data are available at: https://github.com/mushroomcarbon/Spotify."
 date: today
 date-format: long
-abstract: "This paper uses data from Open Data Toronto on TTC subway and bus
-delays to analyze how the delays are generally distributed w.r.t. the time of day. Mean, total, and adjusted statistics were graphed, leading to the conclusions that subways are almost always
-more punctual than buses, peak hour leads to the fewest delays, and leaving 10-15 minutes earlier is a safe margin of error for delays.
-The conclusions of this paper provide a practical, actionable set of guidelines on how citizens can ensure/maximize punctuality when using the TTC to go about the city, as well as potential insight into where resources can be invested in to increase TTC efficiency the most."
+abstract: "This study analyzes the evolution of musical attributes in The National’s discography, focusing on danceability, energy, instrumentalness, and valence across their album releases.
+Using data from Spotify’s API, we analyze how these attributes have fluctuated over the years and visualize their progression through a series of plots.
+Our findings suggest that energy and instrumentalness exhibit more variability, while danceability and valence have followed relatively smoother trajectories.
+These insights contribute to the growing field of music analytics by offering a deeper understanding of the artistic and sonic evolution of a famous band within the indie rock genre."
 format: pdf
@@
 
-The R programming language (@talia), dplyr (@dplyr), opendatatoronto(@opendatatoronto), and tidyverse(@thereferencecanbewhatever) were used to download, modify, and analyse data obtained from Open Data Toronto.
-Styler (@styler) was used to style the code.
-Data used in this paper comes from the "TTC Subway Delays" (@subway_data) and "TTC Bus Delays" (@bus_data) data sets from the Open Data Toronto database, published by the TTC.
+The R programming language [@talia], dplyr [@dplyr], tidyverse[@thereferencecanbewhatever] were used to download, modify, and analyse data obtained from Spotify [@spotify_api] using the SpotifyR package [@spotifyr] regarding music from The National [@the_national].
+Styler [@styler] was used to style the code.
 
@@
 # Load the data
-data <- read_rds(here::here("data/the_national.rds"))
+data <- read_rds(here::here("data/analysis_data/the_national_cleaned.rds"))
 

### 2024-10-10 scripts/01-download_data.R

@@ -12,5 +12,3 @@ library(tidyverse)
 
 the_national <- get_artist_audio_features("the national")
 saveRDS(the_national, "data/the_national.rds")
-
-

### 2024-10-10 scripts/02-data_cleaning.R

@@ -16,11 +16,13 @@ data <- read_rds(here::here("data/raw_data/the_national.rds"))
 
 # Ensure that variables of interest are doubles, removing rows if not
 data <- data %>%
-  filter(is.double(album_release_year),
-         is.double(danceability),
-         is.double(energy),
-         is.double(instrumentalness),
-         is.double(valence))
+  filter(
+    is.double(album_release_year),
+    is.double(danceability),
+    is.double(energy),
+    is.double(instrumentalness),
+    is.double(valence)
+  )
 
 # Save the cleaned data back to the .rds file
 saveRDS(data, here::here("data/analysis_data/the_national_cleaned.rds"))

### 2024-10-10 scripts/03-test_data.R

@@ -23,4 +23,3 @@ is.double(data$danceability)
 is.double(data$energy)
 is.double(data$instrumentalness)
 is.double(data$valence)
-

### 2024-10-10 paper/paper.qmd
llm
paper update discussion

@@
 Styler [@styler] was used to style the code.
+
+Variable Measurements:
+
+Danceability: This metric quantifies how suitable a track is for dancing, ranging from 0.0 to 1.0.
+It considers elements such as tempo, rhythm stability, beat strength, and overall regularity, all influencing a track's likelihood of encouraging dancing.
+
+Energy: Energy is also measured on a scale from 0.0 to 1.0, reflecting the intensity and activity within a track.
+High-energy songs are typically fast-paced, loud, and noisy, while lower-energy tracks are slower and more relaxed.
+Energy is influenced by elements such as dynamic range, perceived loudness, and tempo.
+
+Instrumentalness: This attribute measures the probability of a track being purely instrumental, also ranging from 0.0 to 1.0.
+The closer the value is to 1.0, the greater the likelihood that the track contains no vocals.
+Tracks with values above 0.5 are considered more likely to be instrumental.
+
+Valence: Valence refers to the musical positivity or happiness conveyed by a track, again measured from 0.0 to 1.0.
+Tracks with high valence sound more positive (e.g., happy, cheerful, euphoric), while those with low valence sound more negative or somber (e.g., sad, melancholic).
 
@@
 \FloatBarrier
+
+# Discussion
+
+Spotify’s API uses algorithms to estimate these musical attributes, but like any automated system, these measurements come with inherent limitations:
+
+Danceability and Energy are listed as variable measurements since these metrics are useful, they are influenced by a combination of tempo, rhythm, and loudness, which are subjectively perceived by listeners.
+This can introduce bias, as not all listeners may agree with the "danceability" or "energy" rating that Spotify assigns to a track.
+In addition, The measurement of instrumentals is more straightforward but can still be prone to misclassification, especially in tracks where vocals are used in a non-traditional or minimalistic manner (e.g., vocal samples used as an instrument rather than for lyrics).
+Finally, valence is perhaps the most subjective metric, as it attempts to quantify the emotional tone of a track.
+The algorithm uses musical cues like key, tempo, and mood, but it cannot fully account for lyrical content or the cultural context of the music, which are often significant factors in how a song's emotion is perceived by listeners.
+
+Certain limitations include algorithmic bias, where all the measurements result from machine learning algorithms that may not fully capture the nuanced emotional or structural qualities of music.
+While these algorithms are highly advanced, they are still based on mathematical models that may overlook certain cultural, lyrical, or artistic nuances.
+Another limitation would be the subjectivity of musical interpretation.
+Under this scenario, music is a subjective art form, and attributes like energy or valence are influenced by listener interpretation.
+Thus, there may be discrepancies between what the algorithm detects and what different audiences perceive.
+For example, a song with low valence may still evoke positive emotions depending on the listener's personal experiences.
+Eventually, temporal and contextual variability as Spotify’s measurements are static, meaning they don't account for how the perception of a track might change over time or in different contexts.
+A song may be considered more "danceable" or "happy" depending on where or when it's played, which is not captured in this type of analysis.
+
+By recognizing these limitations, we can more critically engage with the data, understanding that while useful, these metrics provide only one dimension of analysis.
+Future work could incorporate qualitative assessments or crowd-sourced data to cross-check and complement these algorithmically generated insights.
+
+\newpage
+
+# References

### 2024-10-10 paper/paper.qmd
final paper updates

@@
 
-Write Introduction here
+In recent years, the music analytics field has advanced significantly, providing new insights into how artists' sounds evolve over time.
+This study delves into the discography of The National, a prominent indie rock band known for their rich, emotive soundscapes and introspective lyricism, by examining key musical attributes such as danceability, energy, instrumentalness, and valence across their albums.
+By utilizing data drawn from Spotify's API, we aim to track how these characteristics have transformed throughout the band's career, illustrating their progression with comprehensive visualizations.
+Our analysis reveals notable trends and variations, particularly pointing out that energy and instrumentalness exhibit considerable changes, perhaps reflecting shifts in the band’s artistic direction, experimentation, or production choices.
+Conversely, danceability and valence maintain relatively stable trajectories, possibly indicating consistency in their core aesthetic approach.
+These insights not only illuminate The National's evolving sound but also contribute to the broader conversation in music analytics, enhancing our understanding of artistic development and genre dynamics within the indie rock landscape.
+Through this study, we offer a deeper appreciation of how technological tools can be leveraged to explore and comprehend music's intricate evolution over time.
 
@@
 ```
+\FloatBarrier
+
+@fig-dance displays the danceability of music by The National since the band’s establishment.
+The National showed consistently high danceability throughout their career despite a visible drop after their initial 2001 album and their 2017 album.
+This trend reveals that The National had a stable style after it established itself as a rock band.
 
@@
 ```
+\FloatBarrier
+
+Energy was more volatile compared to danceability, as displayed in @fig-energy.
+In the 2001 and 2019 releases, the average energy levels were around 0.5, while in most years, energy levels were greater than 0.6.
+This demonstrates The Nationa’s willingness to experiment, though it held on to its rock identity by producing music with high overall energy.
 
@@
 ```
+\FloatBarrier
+
+Instrumentalness was the most volatile aspect of The National’s musical style over time.
+As shown in @fig-instrumental.
+We can identify that the band experimented with very high instumentalness in 2007 and 2017, while the majority of songs released in other years showed a moderate level of instrumentalness.
+Combined with other trends, we can infer that The National is more willing to attempt music with different compositions, such as songs with more instrumentals and fewer lyrics, but unwilling to shift its fundamental rock style.
 
@@
 \FloatBarrier
+
+Lastly, as indicated in @fig-valence, The Nationals maintained a moderate level of valence throughout the years, consistent with the observation that the band strived to keep its fundamental identity.
 
